From OpenACC to OpenMP 4: \\ Toward Automatic Translation

### 2016-04-29 sig-alternate-sample.tex
Section 3 and start of 4

\newcommand{\squishlist}{
    \begin{list}{$\bullet$}
     { \setlength{\itemsep}{0pt}      \setlength{\parsep}{3pt}
       \setlength{\topsep}{3pt}       \setlength{\partopsep}{0pt}
       \setlength{\leftmargin}{1.5em} \setlength{\labelwidth}{1em}
       \setlength{\labelsep}{0.5em} } }

@@

\section{Translation}
\label{sec:translation}

The major compilers supporting OpenACC are Portland Group (PGI), Cray, and CAPS. On the other hand, Intel has decided not to support OpenACC until accelerators are fully integrated into OpenMP. We describe an automated approach to assist programmers in converting OpenACC to OpenMP, so that same syntax could be used for both Cray and Intel compiler while maintaining the performance. We describe a refactoring (POOD) for this purpose.

\subsection{A Subset of OpenACC}
For the purposes of prototyping, we designed our tool to accept the subset of
OpenACC described by the grammar in Figure~\ref{fig:grammar}.  This includes
all of the directives and clauses used by the EPCC Benchmarks, as well as the
other example programs in our test suite.  While it omits some important
features---asynchronous execution, unstructured data lifetimes, etc.---it does
correspond to a useful, commonly-used subset of OpenACC.

\begin{figure}
\begin{tabular}{rcl}
\textit{acc-directive} &$\rightarrow$& \texttt{\#pragma acc data\ }\textit{data-clauses} \\
                       &     $|$     & \texttt{\#pragma acc parallel}\texttt{\ }\textit{parallel-clauses} \\
                       &     $|$     & \texttt{\#pragma acc loop\ }\textit{type}\texttt{\ }\textit{loop-clauses} \\
\textit{type}          &$\rightarrow$& \texttt{gang} \\
                       &     $|$     & \texttt{vector} \\
                       &     $|$     & \texttt{seq} \\
                       &     $|$     & $\epsilon$ \\
\textit{data-clause}   &$\rightarrow$& \texttt{copyin(}\textit{vars}\texttt{)} \\
                       &     $|$     & \texttt{copyout(}\textit{vars}\texttt{)} \\
                       &     $|$     & \texttt{copy(}\textit{vars}\texttt{)} \\
                       &     $|$     & \texttt{create(}\textit{vars}\texttt{)} \\
\textit{parallel-clause}&$\rightarrow$&\texttt{num\_gangs(}\textit{n}\texttt{)} \\
                       &     $|$     & \texttt{vector\_length(}\textit{n}\texttt{)} \\
                       &     $|$     & \texttt{private(}\textit{vars}\texttt{)} \\
                       &     $|$     & \texttt{reduction(}\textit{reductions}\texttt{)} \\
\textit{loop-clause}   &$\rightarrow$& \texttt{independent} \\
                       &     $|$     & \texttt{private(}\textit{vars}\texttt{)} \\
\end{tabular}
\caption{OpenACC subset grammar accepted by the translation tool.  Syntax is
omitted for \textit{vars}, \textit{reductions}, and integers \textit{n}.  The
nonterminals \textit{data-clauses} and \textit{loop-clauses} correspond to
sequences of zero or more phrases derived from \textit{data-clause} and
\textit{loop-clause}, respectively.  In addition to the directives shown,
\texttt{\#pragma acc parallel loop} is supported, and treated as syntactic
sugar for a \texttt{loop} directive nested directly under a \texttt{parallel}
directive.}
\label{fig:grammar}
\end{figure}

@@

In practice, the situation is not so simple.  The OpenACC and OpenMP
specifications define an \emph{abstract} execution model.  They do not dictate
how that model is mapped to specific hardware devices.  This is left to the
compiler.  Our OpenACC programs were compiled using PGCC~16.1, the standard
commercial compiler for OpenACC\@.  We compiled OpenMP~4 using a custom build
of Clang~3.8\footnote{See \url{https://parallel-computing.pro/index.php/9-cuda/43-openmp-4-0-on-nvidia-cuda-gpus}} (PGCC does not yet support
OpenMP~4\footnote{Cray's compilers support both, but there are currently no
Cray resources in the XSEDE ecosystem.}).  As we will show, directives that
appear similar may be translated quite differently by the two compilers.

\subsubsection{Work Distribution Differs}

\begin{figure*}
\begin{center}
\begin{minipage}{4in}
{\footnotesize
\begin{verbatim}
#pragma acc data copyin(A[0:N], B[0:N]), copyout(C[0:N])
#pragma omp target data map(to:A[0:N], B[0:N]), map(from:C[0:N])
{
    #pragma acc parallel loop
    #pragma omp target teams distribute parallel for
    for (int i = 0; i < N; ++i)
    {
        C[i] = A[i] + B[i];
    }
}
\end{verbatim}
}
\end{minipage}
\end{center}
\caption{Vector addition}
\label{fig:vecadd}
\end{figure*}

Perhaps the simplest example of directive-based accelerator code is the vector
addition kernel shown in Figure~\ref{fig:vecadd}.  It adds two vectors $A$ and
$B$ on an accelerator device, storing the sum in $C$.  While the directives
direct the compiler to parallelize the loop on the accelerator, the compiler
can decide (in this case) how to divide the loop iterations among threads
and thread blocks on the CUDA device.  When $N = 10000$:
\begin{itemize}
\item PGCC generates a grid of 79 thread blocks with 128 threads per block;
kernel execution time is about $3~\mu s$ on a Tesla~K40.
\item Clang generates a grid of 15 thread blocks with 1024 threads per block;
kernel execution time is about $30~\mu s$ on a Tesla~K40.\footnote{Adding
\texttt{num\_teams(79) thread\_limit(128)} to the OpenMP code (to match the
grid and block dimensions of the OpenACC code) actually increases the kernel
execution time to about $80~\mu s$.
%Adding \texttt{num\_gangs(15) vector\_length(1024)} to the OpenACC code does
%not change the performance significantly.
}
\end{itemize}

@@

\begin{figure*}
\begin{minipage}{.5\textwidth}
{\footnotesize
\begin{verbatim}
// OpenACC
#pragma acc parallel loop
for (int i = 0; i < 10; i++) {      // Gang
    for (int j = 0; j < 10; j++) {  // Vector
        ...
    }
}

\end{verbatim}
}
\end{minipage}
\begin{minipage}{.5\textwidth}
{\footnotesize
\begin{verbatim}
// OpenMP 4 Device Directives
#pragma omp target teams distribute
for (int i = 0; i < 10; i++) {
    #pragma omp parallel for
    for (int j = 0; j < 10; j++) {
        ...
    }
}
\end{verbatim}
}
\end{minipage}
\caption{Doubly nested loops.}
\label{fig:nest}
\end{figure*}

More important is the handling of loop nests, such as the simple
doubly nested loops in Figure~\ref{fig:nest}.

In OpenACC, the compiler has some freedom to decide how to parallelize the loop
nest.  PGCC will typically run the outermost loop as a gang loop and the
innermost loop as a vector loop.  Given the OpenACC code in
Figure~\ref{fig:nest}, the compiler does exactly this: the iterations of the
$i$-loop are divided among the thread blocks, and the iterations of the
$j$-loop are divided among the threads within each block.

In OpenMP, work distribution is not so automatic.  Consider these
``obvious'' but incorrect translations of the OpenACC code into OpenMP: The
\texttt{\#pragma acc parallel loop} directive could be replaced by
\begin{itemize}
\item \texttt{\#pragma omp target teams}.  In this case, the entire loop
nest would be run redundantly by all thread blocks.
\item \texttt{\#pragma omp target teams distribute}.  In this case, the
iterations of the $i$-loop would be divided among thread blocks, but only
one thread would be active within each block.
\item \texttt{\#pragma omp target teams distribute parallel for}.  Here,
the iterations of the $i$-loop are distributed among threads and thread blocks.
\end{itemize}
In all three cases, \emph{the inner $j$-loop runs sequentially}, because the
parallelization directive is applied to only one loop.  OpenMP requires the
programmer to explicitly specify how to run each loop in the nest.  The
translation most similar to the OpenACC code is shown on the right in
Figure~\ref{fig:nest}: The iterations of the $i$-loop are divided among thread
blocks using the \texttt{distribute} directive, and the iterations of the
$j$-loop are divided among threads in the block using the \texttt{parallel for}
directive.

@@

In both OpenACC and OpenMP, if \texttt{data}/\texttt{target data} directives
are omitted, the compiler will automatically determine what data needs to be
transferred between the host and the accelerator.  This is another case where
compilers may differ.  In the vector addition example
(Figure~\ref{fig:vecadd}), consider what happens when $A$, $B$, and $C$ are
\emph{pointers} allocated just before the kernel (in the same function) using
\textit{malloc}.  With PGCC, the \texttt{\#pragma acc data} directive can be
omitted; it can determine statically that 10,000 elements of $A$, $B$, and
$C$ are accessed and thus need to be copied.  With Clang, if the
\texttt{\#pragma omp target data} directive is omitted, the kernel crashes at
runtime (\emph{Device Illegal Address}).

\subsection{Translation Algorithm}

\newcommand{\translate}[1]{\ensuremath{\mathcal{T}[\![ #1 ]\!]}}
\begin{figure*}
\begin{align*}
%
\translate{\texttt{\#pragma acc data\ }\textit{data-clauses}} &=
   \texttt{\#pragma omp target data } \translate{\textit{data-clauses}} \\
%
\translate{\texttt{copyin(}\textit{vars}\texttt{)}} &=
   \texttt{map(to:}\textit{vars}\texttt{)} \\
%
\translate{\texttt{copyout(}\textit{vars}\texttt{)}} &=
   \texttt{map(from:}\textit{vars}\texttt{)} \\
%
\translate{\texttt{copy(}\textit{vars}\texttt{)}} &=
   \texttt{map(tofrom:}\textit{vars}\texttt{)} \\
%
\translate{\texttt{create(}\textit{vars}\texttt{)}} &=
   \texttt{map(alloc:}\textit{vars}\texttt{)} \\
%
\translate{\texttt{\#pragma acc parallel\ }\textit{parallel-clauses}} &=
   \left.\parbox[t]{4in}{%
   \texttt{\#pragma omp target teams\ }\translate{\textit{parallel-clauses}}}\right. \\
%
\translate{\texttt{num\_gangs(}\textit{n}\texttt{)}} &=
   \texttt{num\_teams(}\textit{n}\texttt{)} \\
%
\translate{\texttt{vector\_length(}\textit{n}\texttt{)}} &=
   \texttt{thread\_limit(}\textit{n}\texttt{)} \\
%
\translate{\texttt{private(}\textit{vars}\texttt{)}} &=
   \texttt{private(}\textit{vars}\texttt{)} \\
%
\translate{\texttt{reduction(}\textit{vars}\texttt{)}} &=
   \texttt{reduction(}\textit{reductions}\texttt{)} \\
%
\translate{\texttt{\#pragma acc loop\ }\textit{type}\texttt{\ }\textit{loop-clauses}} &=
   \begin{cases}
   \texttt{\#pragma omp distribute\ }\translate{\textit{loop-clauses}} \\
        \hspace*{2em} \textup{ if $\textit{type} \neq \texttt{seq}$, and the loop is the outermost} \\
        \hspace*{3em} \textup{ non-sequential loop in a loop nest} \\
        \hspace*{3em} \textup{ in a \texttt{parallel} region} \vspace*{.25em} \\
   \texttt{\#pragma omp parallel for\ }\translate{\textit{loop-clauses}} \\
        \hspace*{2em} \textup{ if $\textit{type} \neq \texttt{seq}$, and the loop is the innermost} \\
        \hspace*{3em} \textup{ non-sequential loop in a loop nest} \\
        \hspace*{3em} \textup{ in a \texttt{parallel} region} \vspace*{.25em} \\
   \texttt{\#pragma omp distribute parallel for\ }\translate{\textit{loop-clauses}} \\
        \hspace*{2em} \textup{ if both of the above apply} \vspace*{.33em} \\
   \epsilon
        \hspace*{1.65em} \textup{ otherwise} \\
   \end{cases} \\
%
\translate{\texttt{independent}} &=
   \epsilon \\
%
\end{align*}
\caption{Translation rules from OpenACC to OpenMP~4.  The function
$\mathcal{T}$ maps phrases derived from the OpenACC subset grammar to
strings corresponding to one or more OpenMP directives.  Sequences
(e.g., \textit{parallel-clauses}) are translated elementwise, that is,
$\translate{t_1\ t_2\ \dots\ t_n} =
\translate{t_1}\ \translate{t_2}\ \dots\ \translate{t_n}$.}
\label{fig:rules}
\end{figure*}

@@

Our translation algorithm was designed with the following objectives:
\begin{itemize}
\item \emph{The translation should be straightfoward and consistent.}
Each OpenACC directive should translate to the same OpenMP directive(s)
every time, and the translation should mirror the structure of the OpenACC
code, to the extent possible.
\item \emph{Automated translation and manual editing should be complementary.}
The tool should not try to automate too much.  It should not perform loop
nest optimizations; it should not infer implicit data transfers; it should
not make any invasive code changes.  The tool should automate the tedious,
most obvious parts of the translation, in a completely predictable way.
More sophisticated changes---like performance tuning---should be under the
control of the programmer.
\end{itemize}

The translation algorithm is as follows.  The input is a C source file
containing OpenACC directives.  The output is the same file with these
directives replaced by OpenMP directives.
\begin{enumerate}
\item If any of the following occur, warn the user; these structures will
not be automatically converted:
\begin{itemize}
\item The header file \texttt{openacc.h} is included.
\item The macro \texttt{\_OPENACC} is used.
\item Functions in the OpenACC runtime library are called or referenced.
\end{itemize}
\item Label loops using the algorithm described in \S\ref{sec:loops}.
\item Apply the translation rules in Figure~\ref{fig:rules} to convert
OpenACC directives to OpenMP directives.
\end{enumerate}

@@

\subsubsection{Translating Data Directives}

@@

\subsubsection{Translating Loop Nests}
\label{sec:loops}

OpenACC's \texttt{gang} clause is roughly equivalent to OpenMP's
\texttt{distribute}---effectively, it distributes work among thread blocks.
Likewise, OpenACC's \texttt{vector} clause is roughly equivalent to OpenMP's
\texttt{parallel for} inside a \texttt{teams} region: it distributes work among
threads within a block.

Unfortunately, translation isn't quite so simple.  OpenACC does not
\emph{require} the \texttt{gang} and \texttt{vector} clauses to be used, in
which case the work distribution is determined by the compiler.  OpenACC is
also surprisingly lax about where the nesting of gang and vector loops.  For
example, a \texttt{gang} loop may be nested inside a \texttt{vector} loop;
there is no direct equivalent of this in OpenMP.  Moreover, PGCC may ignore the
\texttt{gang} and \texttt{vector} directives altogether, e.g., when a gang loop
appears inside another gang loop, or a vector loop appears outside another
vector loop.

For these reasons, \emph{the loop-handling rule of Figure~\ref{fig:rules}
ignores OpenACC's \texttt{gang} and \texttt{vector} clauses}.  Instead, it uses
a heuristic typical of parallelizing compilers (one that PGCC appears to use,
in fact): the iterations of the outermost parallelizable loop are partitioned
among thread blocks, and the iterations of the innermost parallelizable loop
are partitioned among the CUDA threads within that block.

>>>>

\newcommand{\squish}{
     { \setlength{\itemsep}{0pt}      \setlength{\parsep}{3pt}
       \setlength{\topsep}{3pt}       \setlength{\partopsep}{0pt}
       \setlength{\leftmargin}{1.5em} \setlength{\labelwidth}{1em}
       \setlength{\labelsep}{0.5em} } }

@@

\section{Translation Considerations}
\label{sec:translation}

%The major compilers supporting OpenACC are Portland Group (PGI), Cray, and CAPS. On the other hand, Intel has decided not to support OpenACC until accelerators are fully integrated into OpenMP. We describe an automated approach to assist programmers in converting OpenACC to OpenMP, so that same syntax could be used for both Cray and Intel compiler while maintaining the performance. We describe a refactoring (POOD) for this purpose.

Our goal was to create a tool to convert OpenACC code to equivalent OpenMP
code, automating as much of the process as possible.  For the purposes of
prototyping, we focused on translating C code (as opposed to C++ or Fortran).

\subsection{A Subset of OpenACC}
For the purposes of prototyping, we designed our tool to accept the subset of
OpenACC described by the grammar in Figure~\ref{fig:grammar}.  This includes
all of the directives and clauses used by the example programs in our test
suite (notably, the EPCC Level~1 benchmarks).  While it omits some important
features---asynchronous execution, unstructured data lifetimes, etc.---it does
correspond to a useful, commonly-used subset of OpenACC.

\begin{figure}
\begin{tabular}{rcl}
\textit{acc-directive} &$\rightarrow$& \texttt{\#pragma acc data\ }\textit{data-clauses} \\
                       &     $|$     & \texttt{\#pragma acc parallel}\texttt{\ }\textit{par-clauses} \\
                       &     $|$     & \texttt{\#pragma acc loop\ }\textit{type}\texttt{\ }\textit{loop-clauses} \\
\textit{type}          &$\rightarrow$& \texttt{gang vector} \\
                       &     $|$     & \texttt{gang} \\
                       &     $|$     & \texttt{vector} \\
                       &     $|$     & \texttt{seq} \\
%                       &     $|$     & $\epsilon$ \\
\textit{data-clause}   &$\rightarrow$& \texttt{copyin(}\textit{vars}\texttt{)} \\
                       &     $|$     & \texttt{copyout(}\textit{vars}\texttt{)} \\
                       &     $|$     & \texttt{copy(}\textit{vars}\texttt{)} \\
                       &     $|$     & \texttt{create(}\textit{vars}\texttt{)} \\
\textit{par-clause}&$\rightarrow$&\texttt{num\_gangs(}\textit{n}\texttt{)} \\
                       &     $|$     & \texttt{vector\_length(}\textit{n}\texttt{)} \\
                       &     $|$     & \texttt{private(}\textit{vars}\texttt{)} \\
                       &     $|$     & \texttt{reduction(}\textit{reductions}\texttt{)} \\
\textit{loop-clause}   &$\rightarrow$& \texttt{independent} \\
                       &     $|$     & \texttt{private(}\textit{vars}\texttt{)} \\
\end{tabular}
\caption{OpenACC subset grammar accepted by the translation tool.  Syntax is
omitted for \textit{vars}, \textit{reductions}, and natural numbers \textit{n}.
The nonterminals \textit{data-clauses} and \textit{loop-clauses} correspond to
sequences of zero or more phrases derived from \textit{data-clause} and
\textit{loop-clause}, respectively.  In addition to the directives shown,
\texttt{\#pragma acc parallel loop} is supported, and treated as syntactic
sugar for a \texttt{loop} directive nested directly under a \texttt{parallel}
directive.}
\label{fig:grammar}
\end{figure}

@@

In practice, the situation is not so simple.  The OpenACC and OpenMP
specifications define an \emph{abstract} execution model.  They do not dictate
how that model is mapped to specific hardware devices.  This is left to the
compiler.  Our OpenACC programs were compiled using PGCC~16.1, the standard
commercial compiler for OpenACC\@.  We compiled OpenMP~4 using a custom build
of Clang~3.8~\cite{clang} (since PGCC does not yet support
OpenMP~4).\footnote{Cray's compilers support both, but there are currently no
Cray resources in the XSEDE ecosystem.}  As we will show, directives that
appear similar may be translated quite differently by the two compilers.

\subsubsection{Work Distribution for Single Loops}

\begin{figure}
\begin{center}
\begin{minipage}{4in}
{\footnotesize
\begin{verbatim}
#pragma acc data copyin(A[0:N], B[0:N]), \
                 copyout(C[0:N])
#pragma omp target data map(to:A[0:N], B[0:N]), \
                        map(from:C[0:N])
{
  #pragma acc parallel loop
  #pragma omp target teams distribute parallel for
  for (int i = 0; i < N; ++i)
  {
    C[i] = A[i] + B[i];
  }
}
\end{verbatim}
}
\end{minipage}
\end{center}
\caption{Vector addition kernel.}
\label{fig:vecadd}
\end{figure}

Perhaps the simplest example of directive-based accelerator code is the vector
addition kernel shown in Figure~\ref{fig:vecadd}.  It adds two vectors $A$ and
$B$ on an accelerator device, storing the sum in $C$.  While the directives
direct the compiler to parallelize the loop on the accelerator, the compiler
can decide (in this case) how to divide the loop iterations among threads
and thread blocks on the CUDA device.  When $N = 10000$:
\begin{itemize}\squish
\item PGCC generates a grid of 79 thread blocks with 128 threads per block;
kernel execution time is about $3~\mu s$ on a Tesla~K40.
\item Clang generates a grid of 15 thread blocks with 1024 threads per block;
kernel execution time is about $30~\mu s$ on a Tesla~K40.\footnote{Adding
\texttt{num\_teams(79) thread\_limit(128)} to the OpenMP code (to match the
grid and block dimensions of the OpenACC code) actually increases the kernel
execution time to about $80~\mu s$.
%Adding \texttt{num\_gangs(15) vector\_length(1024)} to the OpenACC code does
%not change the performance significantly.
}
\end{itemize}

@@

\begin{figure*}
\begin{minipage}{.33\textwidth}
\emph{OpenACC (Original):}
{\footnotesize
\begin{verbatim}
#pragma acc parallel loop
for (int i = 0; i < 10; i++) {
  for (int j = 0; j < 10; j++) {
    ...
  }
}

\end{verbatim}
}
\centerline{(a)}
\end{minipage}
\begin{minipage}{.33\textwidth}
\emph{OpenACC (Equivalent):}
{\footnotesize
\begin{verbatim}
#pragma acc parallel loop gang
for (int i = 0; i < 10; i++) {
  #pragma acc parallel loop vector
  for (int j = 0; j < 10; j++) {
    ...
  }
}
\end{verbatim}
}
\centerline{(b)}
\end{minipage}
\begin{minipage}{.33\textwidth}
\emph{OpenMP~4:}
{\footnotesize
\begin{verbatim}
#pragma omp target teams distribute
for (int i = 0; i < 10; i++) {
  #pragma omp parallel for
  for (int j = 0; j < 10; j++) {
    ...
  }
}
\end{verbatim}
}
\centerline{(c)}
\end{minipage}
\caption{Doubly nested loops.}
\label{fig:nest}
\end{figure*}

\subsubsection{Work Distribution for Loop Nests}
\label{sec:nests}

More important is the handling of loop nests, such as the simple
doubly nested loops in Figure~\ref{fig:nest}(a).

In OpenACC, the compiler has some freedom to decide how to parallelize the loop
nest.  PGCC will typically run the outermost loop as a gang loop and the
innermost loop as a vector loop.  Given the OpenACC code in
Figure~\ref{fig:nest}(a), the compiler does exactly this: the iterations of the
$i$-loop are divided among the thread blocks, and the iterations of the
$j$-loop are divided among the threads within each block.  The OpenACC code in
Figure~\ref{fig:nest}(b) is functionally equivalent but makes this distribution
explicit.

In OpenMP, work distribution is not so automatic.  Consider these ``obvious''
but incorrect translations: The \texttt{\#pragma acc parallel loop} directive
could be replaced by:
\begin{itemize}\squish
\item \texttt{\#pragma omp target teams}.  In this case, the entire loop
nest would be run redundantly by all thread blocks.
\item \texttt{\#pragma omp target teams distribute}.  In this case, the
iterations of the $i$-loop would be divided among thread blocks, but only
one thread would be active within each block.
\item \texttt{\#pragma omp target teams distribute parallel for}.  Here,
the iterations of the $i$-loop are distributed among threads and thread blocks.
\end{itemize}
In all three cases, \emph{the inner $j$-loop runs sequentially}, because the
parallelization directive is applied to only one loop.  OpenMP requires the
programmer to explicitly specify how to run each loop in the nest.

The translation most similar to the OpenACC code is shown in
Figure~\ref{fig:nest}(c).  The iterations of the $i$-loop are divided among
thread blocks using the \texttt{distribute} directive, and the iterations of
the $j$-loop are divided among threads in the block using the \texttt{parallel
for} directive.
%
OpenACC's \texttt{gang} clause is roughly equivalent to OpenMP's
\texttt{distribute}---effectively, it distributes work among thread blocks.
Likewise, OpenACC's \texttt{vector} clause is roughly equivalent to OpenMP's
\texttt{parallel for} inside a \texttt{teams} region: it distributes work among
threads within a block.

@@

In both OpenACC and OpenMP, if \texttt{data}/\texttt{target data} directives
are omitted, the compiler will automatically determine what data needs to be
transferred between the host and the accelerator.  This is another case where
compilers may differ.  In the vector addition example
(Figure~\ref{fig:vecadd}), consider what happens when $A$, $B$, and $C$ are
\emph{pointers} allocated just before the kernel (in the same function) using
\textit{malloc}.  With PGCC, the \texttt{\#pragma acc data} directive can be
omitted; it can determine statically that 10,000 elements of $A$, $B$, and
$C$ are accessed and thus need to be copied.  With Clang, if the
\texttt{\#pragma omp target data} directive is omitted, the kernel crashes at
runtime (\emph{device illegal address}).

\section{Translation Algorithm and \\ Tool Implementation}
\label{sec:algorithm}

@@

Therefore, we designed our translation tool---and its underlying
algorithm---with the following objectives:
\begin{itemize}
\item \emph{The translation should be straightfoward and consistent.}
Each OpenACC directive should translate to the same OpenMP directive(s)
every time, and the translation should mirror the structure of the OpenACC
code, to the extent possible.
\item \emph{Automated translation and manual editing should be complementary.}
The tool should not try to automate too much.  It should not perform loop
nest optimizations; it should not infer implicit data transfers; it should
not make any invasive code changes.  The tool should automate the tedious,
most obvious parts of the translation, in a completely predictable way.
More sophisticated changes---like performance tuning---should be under the
control of the programmer.
\end{itemize}

The translation algorithm is as follows.  The input is a C source file
containing OpenACC directives.  The output is the same file with OpenMP
directives substituted.
\begin{enumerate}
\item If any of the following occur, warn the user; these structures will
not be automatically converted:
\begin{itemize}
\item The header file \texttt{openacc.h} is included.
\item The macro \texttt{\_OPENACC} is used.
\item Functions in the OpenACC runtime library are called or referenced.
\end{itemize}
\item Prepare parallel loop nests using the algorithm described in
\S\ref{sec:loops} below.
\item Apply the translation rules described below in \S\ref{sec:directives} to
convert OpenACC directives to OpenMP directives.
\end{enumerate}

\subsection{Loop Nest Preparation}
\label{sec:loops}

As noted in \S\ref{sec:nests} above, OpenACC-to-OpenMP translation is
complicated by the fact that work distribution is explicit in OpenMP, while it
may be left to the compiler in OpenACC\@.  We noted that the code in
Figure~\ref{fig:nest}(a) is equivalent to that in Figure~\ref{fig:nest}(b),
expect the latter makes the work distribution explicit.

Before our translation tool attempts to translate any OpenACC directives,
it ``prepares'' each parallel loop nest as follows.
\begin{itemize}\squish
\item If the loop nest contains explicit gang and vector clauses, there must be
exactly one gang loop and exactly one vector loop in the nest.  The same loop
may be both the gang and vector loop.
\item If the loop nest does not contain an explicit gang loop, the outermost
non-sequential loop is chosen as the gang loop.
\item If the loop nest does not contain an explicit vector loop, the innermost
non-sequential loop is chosen as the vector loop.
\item All loops in the loop nest other than the gang and vector loop are run
sequentially.
\item If a loop will be run sequentially, and it contains a \texttt{private}
clause, all \texttt{private} variables must not be live at the top of the loop.
\item If the loop nest does not have the form above (e.g., if it contains two
gang loops, or if a vector loop appears outside a gang loop), the translation
to OpenMP cannot proceed.
\end{itemize}

The last rule is necessary because OpenACC is surprisingly lax about where the
nesting of gang and vector loops.  For example, a \texttt{gang} loop may be
nested inside a \texttt{vector} loop; there is no direct equivalent of this in
OpenMP.  Moreover, PGCC may ignore the \texttt{gang} and \texttt{vector}
directives altogether, e.g., when a gang loop appears inside another gang loop,
or a vector loop appears outside another vector loop.

The next-to-last rule ensures that, if a loop is run sequentially, the
\texttt{private} clause can be removed without changing the behavior of the
loop.  Although the rule is stated in terms of a live variables
analysis~\cite{kennedy}, the concept is straightforward.  If a variable $a$
is declared as private, but each iteration assigns a value to $a$ before
reading its value, then the \texttt{private} clause can be eliminated if
the loop is run sequentially.  It cannot be removed if each iteration of the
loop depends on the value of $a$ prior to the loop (e.g., if the first
instruction in the loop was \texttt{a++}).

The second and third rules implement a heuristic typical of parallelizing
compilers (one that PGCC appears to use, in fact): the iterations of the
outermost parallelizable loop are partitioned among thread blocks, and the
iterations of the innermost parallelizable loop are partitioned among the CUDA
threads within that block.

To determine if a loop is parallelizable (for the second and third rules), our
tool uses a combination of syntactic checks and dependence
analysis~\cite{wolfe,kennedy}:
\begin{enumerate}\squish
\item If the loop contains a \texttt{seq} clause, it is not parallelizable.
\item If the loop contains a \texttt{gang} or \texttt{vector} clause, it is
parallelizable.
\item Otherwise, we perform a dependence analysis on the loop.  If the loop
does not carry a dependence, it is parallelizable~\cite{kennedy}; if it
does, we determine it to be non-parallelizable.
\end{enumerate}

\subsection{Directive Translation}
\label{sec:directives}

\newcommand{\translate}[1]{\ensuremath{\mathcal{T}[\![ #1 ]\!]}}
\begin{figure*}
\begin{align*}
%
\translate{\texttt{\#pragma acc data\ }\textit{data-clauses}} &=
   \texttt{\#pragma omp target data } \translate{\textit{data-clauses}} \\
%
\translate{\texttt{copyin(}\textit{vars}\texttt{)}} &=
   \texttt{map(to:}\textit{vars}\texttt{)} \\
%
\translate{\texttt{copyout(}\textit{vars}\texttt{)}} &=
   \texttt{map(from:}\textit{vars}\texttt{)} \\
%
\translate{\texttt{copy(}\textit{vars}\texttt{)}} &=
   \texttt{map(tofrom:}\textit{vars}\texttt{)} \\
%
\translate{\texttt{create(}\textit{vars}\texttt{)}} &=
   \texttt{map(alloc:}\textit{vars}\texttt{)} \\
%
\translate{\texttt{\#pragma acc parallel\ }\textit{par-clauses}} &=
   \left.\parbox[t]{4in}{%
   \texttt{\#pragma omp target teams\ }\translate{\textit{par-clauses}}}\right. \\
%
\translate{\texttt{num\_gangs(}\textit{n}\texttt{)}} &=
   \texttt{num\_teams(}\textit{n}\texttt{)} \\
%
\translate{\texttt{vector\_length(}\textit{n}\texttt{)}} &=
   \texttt{thread\_limit(}\textit{n}\texttt{)} \\
%
\translate{\texttt{private(}\textit{vars}\texttt{)}} &=
   \texttt{private(}\textit{vars}\texttt{)} \\
%
\translate{\texttt{reduction(}\textit{vars}\texttt{)}} &=
   \texttt{reduction(}\textit{reductions}\texttt{)} \\
%
\translate{\texttt{\#pragma acc loop\ }\textit{type}\texttt{\ }\textit{loop-clauses}} &=
   \begin{cases}
   \texttt{\#pragma omp distribute parallel for\ }\translate{\textit{loop-clauses}} \\
        \hspace*{2em} \textup{ if $\textit{type} = \texttt{gang vector}$} \vspace*{.25em} \\
   \texttt{\#pragma omp distribute\ }\translate{\textit{loop-clauses}} \\
        \hspace*{2em} \textup{ if $\textit{type} = \texttt{gang}$} \vspace*{.25em} \\
   \texttt{\#pragma omp parallel for\ }\translate{\textit{loop-clauses}} \\
        \hspace*{2em} \textup{ if $\textit{type} = \texttt{vector}$} \vspace*{.25em} \\
   \epsilon
        \hspace*{1.65em} \textup{ if $\textit{type} = \texttt{seq}$} \\
   \end{cases} \\
%
\translate{\texttt{independent}} &=
   \epsilon \\
%
\end{align*}
\caption{Translation rules from OpenACC to OpenMP~4.  The function
$\mathcal{T}$ maps phrases derived from the OpenACC subset grammar to
strings corresponding to one or more OpenMP directives.  Sequences
(e.g., \textit{par-clauses}) are translated elementwise, that is,
$\translate{t_1\ t_2\ \dots\ t_n} =
\translate{t_1}\ \translate{t_2}\ \dots\ \translate{t_n}$.}
The symbol $\epsilon$ denotes the empty string.
\label{fig:rules}
\end{figure*}

The loop nest preparation algorithm in the previous section identifies a single
gang loop and a single vector loop in each parallel loop nest, and it guarantees
that the vector loop is nested under (or is the same as) the gang loop.  After
gang, vector, and sequential loops have been identified, it is possible to
proceed with translating OpenACC directives to OpenMP directives.

@@

@@

Also, these rules assume that each loop in a parallel loop nest has been
\emph{explicitly} labeled with a \texttt{gang}, \texttt{vector}, or
\texttt{seq} clause (or both \texttt{gang} and \texttt{vector}).  Of course,
this is not necessary.  In fact, our tool runs the loop nest preparation
algorithm implicitly; it identifies a gang and vector loop in each parallel
loop nest, but it does not literally insert OpenACC \texttt{gang},
\texttt{vector}, and \texttt{seq} clauses into the source code (although
it retains them if they are already present in the code).
